The evolution of gene expression in the term placenta of viviparous mammals

### 2015-10-13 manuscript/manuscript.Rnw
discuss more completely the transcription profile in homo sapiens

The top 50 expressed genes as measured by \ac{FPKM} in the term
placenta of \textit{Homo sapiens} are shown in
\fref{tab:top50_human_placenta} with a list of all genes in
Supplemental Table X. The gene with by far the highest \ac{FPKM} value
(\Sexpr{gene.expression(species.name="homo sapiens",gene="CSH1")}) %
in humans is \ac{CSH1}, followed by \ac{CSH2}, and \ac{CSHL1}, all
three of which are members of a distinct expansion of the growth
hormone gene family within hominoids. Another member of the growth
hormone gene family, \ac{GH2} (\ac{FPKM}=%
\Sexpr{gene.expression(species.name="homo sapiens",gene="GH2")}), %
is also expressed to a high degree in the human placenta
(\fref{tab:top50_human_placenta}).
% 
Ten pregnancy-specific beta-1-glycoprotein genes are highly-expressed
(\ac{FPKM}>100) in the term placenta of humans, with \ac{PSG4} showing the
greatest degree of expression %
(\ac{FPKM}=\Sexpr{gene.expression(species.name="homo sapiens",gene="PSG4")}). % 
These
genes are members of a hominoid-specific expansion of the
carcinoembryonic antigen (CEA) family and known to be expressed by the
syncytiotrophoblast (Lei et al., 1992; Rand and Cross, 2008). Genes
involved in the production of estrogen (CYP19A1) and progesterone
(HSD3B1) show exceptionally high expression in the human placenta
(\ac{FPKM} = 4817.15 and 829.846 respectively), as well as CGA
(\ac{FPKM}= 4793.08), which generates the alpha subnit of multiple
human chorionic gonadotropin hormones (Table 2). The genes producing
the beta subunits of the chorionic gonadotropins (CGB, CGB1, CGB2,
CGB5, CGB7, CGB8) are either not expressed (CGB8) or expressed
modestly (FPKM≤20.826; Table 8). Other highly expressed genes in the
top 50 include members of the hemoglobin family (e.g. HBA1, HBB, HBG2,
HBA2), the maternally imprinted long noncoding RNA gene H19, genes
coding for proteins required for growth and/or cell proliferation
(ADAM12, KISS1, IGF2, GPC3, CSF3R), anticoagulation (TFPI2), and
glucose transport (SLC2A1) among others (Table 2, Supplemental Table
X)

>>>>

<<cgb.min.expression,results="hide",echo=FALSE,warnings=FALSE>>=
cgb.min.expression <- 
    min(sapply(paste0("CGB",c("","1","2","5","7","8")),
               function(x){gene.expression(species.name="homo sapiens",gene=x)}))
cgb.sum.expression <- 
    sum(sapply(paste0("CGB",c("","1","2","5","7","8")),
               function(x){gene.expression(species.name="homo sapiens",gene=x)}))
@

The top 50 expressed genes as measured by \ac{FPKM} in the term
placenta of \textit{Homo sapiens} are shown in
\fref{tab:top50_human_placenta} with all genes with \ac{FPKM} $\ge 50$
in Supplemental \fref{tab:all_human_placenta}. The gene with by far
the highest \ac{FPKM} value
%
(\Sexpr{gene.expression(species.name="homo sapiens",gene="CSH1")}) %
%
in humans is \textit{CSH1}, followed by \textit{CSH2}, and \textit{CSHL1}, all
three of which are members of a distinct expansion of the growth
hormone gene family within hominoids (~\ref{fig:galectins_evolution_tree}). Another member of the growth
hormone gene family, \ac{GH2} (\ac{FPKM}=%
\Sexpr{gene.expression(species.name="homo sapiens",gene="GH2")}), %
is also expressed to a high degree in the human placenta
(\fref{tab:top50_human_placenta}).
% 
Eleven pregnancy-specific beta-1-glycoprotein genes are highly-expressed
(\ac{FPKM}$\ge 100$) in the term placenta of humans, with \textit{PSG1} showing the
greatest degree of expression %
(\ac{FPKM}=\Sexpr{gene.expression(species.name="homo sapiens",gene="PSG1")}) % 
and the pseudogene \textit{PSG10P} showing the least %
(\ac{FPKM}=\Sexpr{gene.expression(species.name="homo sapiens",gene="PSG10P")}, \fref{fig:psg_evolution_tree}) . %
These
genes are members of a hominoid-specific expansion of the \ac{CEA} family
and known to be expressed by the syncytiotrophoblast
\cite{Rawn.ea2008:evolutionregulationandfunction,Lei.ea1992:Cloningandexpressionof}. %
%
Genes involved in the production of estrogens (%
\Sexpr{gene.expression(species.name="homo sapiens",gene="CYP19A1",complete=TRUE)},
% 
\Sexpr{gene.expression(species.name="homo sapiens",gene="HSD17B1",complete=TRUE)}, and
% 
\Sexpr{gene.expression(species.name="homo sapiens",gene="HSD3B1",complete=TRUE)}%
% 
) and progesterone (%
\Sexpr{gene.expression(species.name="homo sapiens",gene="HSD3B1",complete=TRUE)},
\Sexpr{gene.expression(species.name="homo sapiens",gene="CYP11A1",complete=TRUE)}, and
\Sexpr{gene.expression(species.name="homo sapiens",gene="ADM",complete=TRUE)})
%
show exceptionally high expression in the human placenta, as well as %
%
\textit{CGA} (FPKM=\Sexpr{gene.expression(species.name="homo sapiens",gene="CGA")}, %
%
which generates the alpha subnit of multiple human chorionic
gonadotropin hormones (\fref{tab:top50_human_placenta}). The genes
producing the beta subunits of the chorionic gonadotropins
(\textit{CGB}, \textit{CGB1}, \textit{CGB2}, \textit{CGB5},
\textit{CGB7}, \textit{CGB8}) are either not expressed (%
%
\Sexpr{gene.expression(species.name="homo sapiens",gene="CGB8")}
%
) have reasonable expression ($\mathrm{FPKM} \ge \Sexpr{cgb.min.expression}$),
and the sum of their expression is on the order of the expression of
expression of \textit{CGA}
($\sum \mathit{CGBs}=\Sexpr{cgb.sum.expression}$ FPKM) .
%
%
Other highly expressed genes in the top 50 include members of the
hemoglobin family (e.g. \textit{HBA1}, \textit{HBB}, \textit{HBG2},
\textit{HBA2}), the maternally imprinted long noncoding RNA gene
\textit{H19}, genes coding for proteins required for growth and/or
cell proliferation (\textit{ADAM12}, \textit{KISS1}, \textit{IGF2},
\textit{GPC3}, \textit{CSF3R}), anticoagulation (\textit{TFPI2}), and
glucose transport (\textit{SLC2A1}) among others
(\fref{tab:top50_human_placenta}, Supplemental
\fref{tab:all_human_placenta}).

\begin{figure}
<<psg.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=

tree.id <-
    genes.to.tree[combined.fpkm[gene_short_name=="PSG1" & species=="homo sapiens",gene_id],tree]
gene.tree.table <- trees[[tree.id]]$table
gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression,min.fpkm=5)
@ 
\caption{Expression of pregnancy-specific glycoproteins}
\label{fig:psg_evolution_tree}
\end{figure}
